---
title: "Paper Political Behavior"
format: html
editor: visual
---

```{r}

```

Pedro Abelin

Pedro - GVPT
771: Seminar in American Political Behavior

**The Nature of Belief Systems in Mass Publics: a theory of elitism**

**Introduction**

The pieces of week four present an essential discussion about ideology
(or belief systems) and reasoning and how citizens make sense of reality. In
this paper, I focus on the seminal paper of Converse (1964) and present a
methodological and substantive critique of the work. I contend that the dualism
of the conservative-liberal continuum cannot encompass the complexity of the
social fabric. Furthermore, Converse has a narrow conception of ideology, which
directly influences the conclusions drawn from the survey. Finally, I will
argue that his conclusions reinforce the elitist thesis of democratic theory.

**The Nature of Belief Systems in Mass Publics: core argument**

Using survey data from 1956, 1958, and 1960, Converse (1964) analyzes
and measures ideologies (the author prefers to call "belief systems) and
party identification in the United States. The author defines a belief system
"as a configuration of ideas and attitudes in which the elements are bound
together by some form of constraint or functional interdependence" (1964,
p. 3). In this sense, the author claim that "constraint" is
understood as the extent to which we would successfully predict that a person
based upfront knowledge that an individual has a specified attitude, that he or
she has certain additional ideas and attitudes (Idem). Converse found, however,
that the population does not elaborate on their opinions and make sense of
reality in an ideological sense. For instance, only 10% of people could even
recognize and understand what terms like "conservatism" and "liberalism"
meant (Converse, 1964. 25). That is, people do not think consistently, lacking
a sophisticated ideological structure, where ideology does not play a
significant role in the way people interprets events and shapes their opinions.
In this sense, the liberal-conservative continuum -- critical to the American
political discussion -- is an abstraction of a high order, which are not a
typical conceptual tool for ordinary people (1964, p. 13).

The elites, on the
other side, have different kinds and levels of constraints across belief
elements (Converse, 1964, p. 32), and a highly-structured belief system. In
other words, while the masses have a volatile belief system, the elites
demonstrate a predictable and sophisticated system. The masses, poorly educated,
would be relegated to thinking about more "concrete" things related
to their immediate world - like their family budget - while the elites would
have the capacity to think in more abstract ways, such as many policy-oriented
issues (Converse, 1964, p. 11).

**Conservative-Liberal continuum: a real consensus?**

What does it mean to be a liberal or a conservative? Although (mainly)
American Political Science uses "liberalism" as a synonym for
progressive politics, the truth is that scholars may be using these
terminologies less rigorously than we imagine. Would classical thinkers of
political thought that we consider liberal, for example, be considered liberals
within the liberal-conservative continuum? For example, John Locke is
considered the "father" of liberalism. Locke was enthusiastic of
liberty and freedom since these liberties extend to the limits of the
proprietors' interests (Locke, 1967). The classic liberal was the
author of Carolina's absolute proprietary constitution in 1669, a document that
sanctioned transatlantic slavery (Farr, 2020; Losurdo, 2014). In the same vein,
we can think of the liberalism of the Founding Fathers and their relationship
to slavery. For the Federalists, the people were invoked as a political
category in favor of less democratic principles, as part of the content of the
US constitution was to attenuate popular power in defense of private property
(Wood, 2007, p. 192). Stuart Mill, the most influential liberal of the 19th
century, although an advocate of suffrage for women and workers, was a defender
of the "civilizational drive" of colonialism (Arneil, 1996; Bell,
2010; Willians, 2021). These examples show that the divide between
liberal/conservative may not make sense and that ideologies such as
conservatism and liberalism can have the same philosophical matrix. Or, the
reality can be more complex than our categories.

This long introduction is not by chance. My goal is to demonstrate that
1) the concepts of liberalism/conservatism are more multifaceted than we
imagine 2) if the political theory itself and "experts" also use them
in a diverse and sometimes non-rigorous way, why should we demand the masses
have total clarity about what these terms are supposed to mean -- or the elites
want to mean? However, I intend to go further in my critique: is it possible to
gauge voting expertise by understanding terms such as liberal and conservative?
Or rather, is complete comprehension of these terms a prerequisite for the
voter's understanding of political events and phenomena?

The elitist conclusions of Converse have many consequences. One of the
leading direct consequences of Converse's finding is to assume that both the
grammar and the mosaic of signs and meanings of the masses must be subordinate
to the ones of the elites. First, one must ask why the use and recognition of
Liberal and Conservative is used to evaluate political sophistication and competence.
Second, ideology is a matter of "discourse" and not of
"language" (Eagleton, 1993, p. 9). As Eagleton (Idem) argues, it is
impossible to determine whether a statement is ideological in isolation from
its discursive context. The same piece of language and the terms used by people
can be ideological in one context, not in another; ideology is a function of
the relation of an enunciation to its social context (Eagleton, 1993, p. 10).
The survey results, ironically, may show more consistency in people's rationality
than we think. This goes in the direction of the literature review
provided by Jost et al. (2009). That is, even if people are not perfect in
using abstract ideological concepts, the majority of the public can and do
utilize a subset of fundamental values or principles that can be regarded as
ideological (Jost et al., 2009, p. 311). In a similar direction, Sniderman et
al. (1991) argue that people can, even if only temporarily, strengthen the ties
between political ideas.

Moreover, the apparent incoherence in the reasoning of
the masses may indicate that the social fabric and masses' worldviews are more
complex than Political Science models can account for. This multifaceted and
complex reality is closely connected with what Converse understands as
ideology. Although Converse prefers to use "belief systems," he tends
to equalize the concepts. In this sense, Converse understands ideology as a
simple set of abstract ideas or logical rules. However, ideologies, as Dias
shows (1995, p. 27), are multiple and contradictory because they are
rationalities of various classes at different times and conjunctures. In this
sense, the social interests, which are historically positioned and articulated
in a conflicting way, manifest themselves in the various and conflicting
ideological discourses. Ideologies are a product of a specific time, and the
mosaic of supposed inconsistencies that Converse found may be the shape in
which certain ideologies were manifesting at that time. The American social
fabric has changed a lot in the last decades, so a survey in the same
parameters would probably find much less "inconsistency" among the
representatives. This is not to suggest that people have become ideologically
more sophisticated, but rather that the current contingencies are the result of
a new historical process and that new kinds of antagonism politically produce
the ontology of the social. The evidence brought by the growing literature on
polarization is an example of this reality.

Understanding that ideology is best explained by
individuals' attributes (or lack thereof) exposes Converse's methodological
individualism. Having people move their support from one side of the
ideological spectrum to the other does not meet the expectations of a model
premised on the conception of individuals with predefined preferences and
constraints. In this sense, his methodology faces the limits to understanding
the nature and logic of forming collective identities. Converse fails to
capture a deep sense of morality, the logic and feelings that move those we try
to understand. Participating (or not participating) in politics is not just a
matter of pure information or the pursuit of individual interest. In order to
understand the supposed inconsistency in people's reasoning, it is necessary to
delve into the study of moral values, of the perception that political subjects
have of reality, and to go deep into people's identities and representations.
Such a strong statement that the masses, unlike the elites, have inconsistent
and unsophisticated ideologies has significant consequences for democratic
theory, as we will discuss in the concluding remarks. This urges a deeper study
of people's values, which would require additional methods. In this sense,
Converse would be benefited from the use of the focal groups' research
technique. As a complementary technique, the focal groups could use the
interactions between people to collect information about people's beliefs and
perceptions of politics. This would allow Converse to understand better how
people make sense of their realities. In addition, as Achen (1975) points out,
Converse's own methodology has serious flaws. As he demonstrates (1975, p.
1221), the respondent response variation shows Converse observational errors.
The variability within the survey questions is the vagueness with which the
subjacent attitude is reflected by the survey instrument -- when the survey
questions are vague, entirely stable respondents will appear to be inconsistent
(Idem). In this sense, the inconsistencies may reside in the questions
themselves.

**Final remarks: hypothesis and consequences to democratic theory**

Converse relies on a notion of human nature that condemns the masses to total political ignorance. In this sense, he presents a very problematic
hypothesis: the population is uninformed, irrational, selfish, and unable to
govern itself. Most people would not be able to govern because they would
decide irrationally and be easily manipulated by the \"political\". In this view,
it is not enough to change institutions. After all, disinformation is found in
individuals, not institutions (Miguel, 2002, p. 500). It is important to note
how the notion of inequality is naturalized in this thesis - in which a
dominated mass and the selected ones constitute the political elites. The
selected groups would escape the mediocrity of the masses, destined to be
dominated. In this sense, I argue that Converse's work contributes to the
construction of an elitist hegemony in Political Science.

**References**

Achen, C. H. (1975). Mass political attitudes and the survey response. American
Political Science Review, 69(4), 1218-1231.

Arneil, B. (1996). John Locke and America: the defence of
English colonialism. Oxford University Press.

Bell, D. (2010). John Stuart Mill on Colonies. Political
theory, 38(1), 34-64.

Converse Philip, E. (1964). The Nature of Belief Systems in Mass Publics
Ideology and Discontent David E. Apter New York Free Press, 206, 61.

DIAS, E. (1995). Cidadania e racionalidade de classe. Universidade e
Sociedade. Andes, São Paulo, (11).

Eagleton, T. (1993). Ideology: an introduction.

Farr, J. (2020). " Absolute Power and Authority": John Locke
and the Revisions of The Fundamental Constitutions of Carolina. Locke
Studies, 20, 1-49.

Jost, John T., Christopher M. Federico, and Jaime L. Napier. 2009.
\"Political ideology: Its structure, functions, and elective affinities.\" Annual
review of psychology 60: 307-337.

Losurdo, D. (2014). Liberalism: A counter-history. Verso Books.

Miguel, L. F. (2002). A democracia domesticada: bases antidemocráticas
do pensamento democrático contemporâneo. Dados, 45, 483-511.

Sniderman, P. M., Brody, R. A., & Tetlock, P. E. (1993). Reasoning
and choice: Explorations in political psychology. Cambridge University Press.

Williams, D. (2021). John Stuart Mill and the practice of colonial rule
in India. Journal of International Political Theory, 17(3), 412-428.

Wood, E. M. (2007). Democracy against capitalism; Renewing
historical materialism. Aakar Books.

```{r}

```
